---
title: "Dissertation themes"
author: "K. Bret Staudt Willet"
date: "7/27/2020"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(hrbrthemes)
library(viridis)
library(seriation)

themes <- read_csv("dissertation-code-document-table.csv")
modalities <- read_csv("dissertation-code-cooccurrence.csv")
```

```{r, include=FALSE}
themes_data <-
        themes %>% 
        column_to_rownames('X1')
names(themes_data) <- 
        c("Amelia", "Anne",  "Blair",  "Hallie", "Julie", 
          "Mike", "Simone", "Taylor",  "Wallace")
row.names(themes_data) <- 
        c("Modality: Offline", "Modality: Online grad school", "Modality: Social media",
          "Reasons: School resources", "Reasons: Change", 
          "Reasons: Personal-professional boundaries", "Reasons: Teacher capital",
          "Reasons: Two-worlds pitfall", "Reasons: Connecting", "Reasons: Enacting practice",
          "Reasons: Learning through observation", "Reasons: Planning",
          "Supports: Assigned and appreciated", "Supports: Assigned but unhelpful",
          "Supports: Latent",  "Supports: Self-directed",
          "Relationships: In-school", "Relationships: Out-of-school",
          "Engagement: Information consumer",  "Engagement: Self-seeking contributor",
          "Engagement: Vocationalist"
          )
themes_matrix <- as.matrix(themes_data)

modalities_data <-
        modalities %>% 
        column_to_rownames('X1')
names(modalities_data) <- c("Offline", "Online Grad School", "Social Media")
row.names(modalities_data) <- 
        c("Reasons: School resources", "Reasons: Change", 
          "Reasons: Personal-professional boundaries", "Reasons: Teacher capital",
          "Reasons: Two-worlds pitfall", "Reasons: Connecting", "Reasons: Enacting practice",
          "Reasons: Learning through observation", "Reasons: Planning",
          "Supports: Assigned and appreciated", "Supports: Assigned but unhelpful",
          "Supports: Latent",  "Supports: Self-directed",
          "Relationships: In-school", "Relationships: Out-of-school",
          "Engagement: Information consumer",  "Engagement: Self-seeking contributor",
          "Engagement: Vocationalist"
          )
modalities_matrix <- as.matrix(modalities_data)
```

## Heatmap plots of themes from interviews with new teachers

These plots provide a very quick and rough look at clusters of new teachers and self-reported themes from interviews using the `heatmap()` function from {base} R and `geom_tile()` from {ggplot2}.

### Heatmap in {base} R, Viridis colors, Scaled but Unordered

```{r, echo=FALSE,  fig.width=16, fig.height=9}
heatmap(themes_matrix, Rowv = NA, Colv = NA, scale = 'column',col = viridis(256))
```

### Heatmap in {base} R, Viridis colors, Scaled and Computationally-sorted

```{r, echo=FALSE, fig.width=16, fig.height=9}
heatmap(themes_matrix, scale = 'column',col = viridis(256))
```

### Heatmap in {base} R, Heat colors, Scaled and Computationally-sorted

```{r, echo=FALSE, fig.width=16, fig.height=9}
heatmap(themes_matrix, scale = 'column', col = heat.colors(256))
```

### Themes Heatmap in {ggplot2}, Viridis colors, Scaled by max value, but Unordered

```{r, echo=FALSE, fig.width=16, fig.height=9}
rescale_themes <- function(x) {x * 20 / max(x, na.rm = TRUE)}

themes_to_plot_unordered <-
        themes_data %>%
        rownames_to_column(var = "theme") %>%
        mutate(across(Amelia:Wallace, rescale_themes)) %>%
        pivot_longer(cols = Amelia:Wallace, 
                     names_to = "name")

ggplot(themes_to_plot_unordered, aes(x = name, y = theme)) +
        geom_tile(aes(fill = value), color = 'white', size = 0.5) + 
        scale_fill_viridis(discrete = FALSE) +
        #coord_fixed(ratio = .4) +
        xlab("Participant") +
        ylab("Code") +
        scale_y_discrete(position='right') +
        theme_bw() +
        theme(axis.title = element_text(size=32, family='serif'),
              axis.text = element_text(size=18, family='serif'),
              legend.title=element_text(size=18, family='serif'), 
              legend.text=element_text(size=12, family='serif'),
              legend.position='left',
              panel.background=element_blank(),
              panel.border=element_blank(),
              panel.grid.major=element_blank(),
              panel.grid.minor=element_blank(),
              plot.background=element_blank()
        ) +
        labs(fill = "Scaled Value")
```

### Themes Heatmap in {ggplot2}, Viridis colors, Scaled by max value, and Computationally-sorted 

This final plot uses principal component analysis (PCA) through the {seriation} package to sort and cluster the heatmap both by most-similar participants and related themes. 

```{r, echo=FALSE, fig.width=16, fig.height=9}
themes_scaled_ordered <-
        themes_data %>%
        rownames_to_column(var = "theme") %>%
        mutate(across(Amelia:Wallace, rescale_themes)) %>%
        column_to_rownames('theme') %>%
        as.matrix() %>%
        seriate()

new_levels_scaled_theme <- row.names(themes_matrix)[get_order(themes_scaled_ordered, dim=1)]
new_levels_scaled_name <- colnames(themes_matrix)[get_order(themes_scaled_ordered, dim=2)]

themes_to_plot_ordered <- 
        themes_to_plot_unordered
themes_to_plot_ordered$theme <- 
        factor(themes_to_plot_ordered$theme, levels = rev(new_levels_scaled_theme))
themes_to_plot_ordered$name <- 
        factor(themes_to_plot_ordered$name, levels = rev(new_levels_scaled_name))

ggplot(themes_to_plot_ordered, aes(x = name, y = theme)) +
        geom_tile(aes(fill = value), color = 'white', size = 0.5) + 
        scale_fill_viridis(discrete = FALSE) +
        #coord_fixed(ratio = .4) +
        xlab("Participant") +
        ylab("Code") +
        scale_y_discrete(position='right') +
        theme_bw() +
        theme(axis.title = element_text(size=32, family='serif'),
              axis.text = element_text(size=18, family='serif'),
              legend.title=element_text(size=18, family='serif'), 
              legend.text=element_text(size=12, family='serif'),
              legend.position='left',
              panel.background=element_blank(),
              panel.border=element_blank(),
              panel.grid.major=element_blank(),
              panel.grid.minor=element_blank(),
              plot.background=element_blank()
        ) +
        labs(fill = "Scaled Value")
```

```{r, include=FALSE, eval=FALSE}
ggsave("dissertation-themes-heatmap-sorted.png", width = 16, height = 9)
```

### Themes Heatmap in {ggplot2}, Viridis colors, Scaled by percent presence, and Computationally-sorted 

```{r, echo=FALSE, fig.width=16, fig.height=9}
themes_quote_count <- c(36, 35, 26, 35, 38, 28, 34, 26, 38)
themes_scaled <- themes_matrix
for(i in 1:9) {
       themes_scaled[, i] <- 
               100 * themes_scaled[, i] / themes_quote_count[i]
}

themes_scaled_ordered2 <- seriate(themes_scaled)

new_levels_scaled_theme2 <- row.names(themes_matrix)[get_order(themes_scaled_ordered2, dim=1)]
new_levels_scaled_name2 <- colnames(themes_matrix)[get_order(themes_scaled_ordered2, dim=2)]

themes_to_plot_ordered2 <- 
        themes_scaled %>%
        as_tibble() %>%
        mutate(theme = row.names(themes_scaled)) %>%
        select(theme, everything()) %>%
        pivot_longer(cols = Amelia:Wallace, 
                     names_to = "name")
themes_to_plot_ordered2$theme <- 
        factor(themes_to_plot_ordered2$theme, levels = rev(new_levels_scaled_theme2))
themes_to_plot_ordered2$name <- 
        factor(themes_to_plot_ordered2$name, levels = rev(new_levels_scaled_name2))

ggplot(themes_to_plot_ordered2, aes(x = name, y = theme)) +
        geom_tile(aes(fill = value), color = 'white', size = 0.5) + 
        scale_fill_viridis(rescaler = function(x, to = c(0, 1), from = NULL) {
                ifelse(x < 50, 
                       scales::rescale(x,
                                       to = to,
                                       from = c(min(x, na.rm = TRUE), 50)),
                       1)
        }
        ) +
        xlab("Participant") +
        scale_x_discrete(position='top') +
        ylab("Thematic Code") +
        scale_y_discrete(position='right') +
        theme_bw() +
        theme(axis.title = element_text(size=32, family='serif'),
              axis.text = element_text(size=18, family='serif'),
              legend.title = element_text(size=18, family='serif'), 
              legend.text = element_text(size=12, family='serif'),
              legend.position='bottom',
              legend.background = element_blank(),
              legend.box.background = element_rect(colour='black'),
              panel.background = element_blank(),
              panel.border = element_blank(),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              plot.background = element_blank()
        ) +
        labs(fill = "Percentage of Quotations")
```

```{r, include=FALSE, eval=FALSE}
ggsave("dissertation-themes-heatmap-sorted2.png", width = 16, height = 9)
```

### Modalities Heatmap in {ggplot2}, Viridis colors, Scaled but Unordered

```{r, echo=FALSE, fig.width=16, fig.height=9}
rescale_modalities <- function(x) {x * 20 / max(x, na.rm = TRUE)}

modalities_to_plot_unordered <-
        modalities_data %>%
        rownames_to_column(var = "theme") %>%
        mutate(across(Offline:`Social Media`, rescale_modalities)) %>%
        pivot_longer(cols = Offline:`Social Media`, 
                     names_to = "modality")

ggplot(modalities_to_plot_unordered, aes(x = modality, y = theme)) +
        geom_tile(aes(fill = value), color = 'white', size = 0.5) + 
        scale_fill_viridis(discrete = FALSE) +
        #coord_fixed(ratio = .4) +
        xlab("Modality") +
        ylab("Theme") +
        scale_y_discrete(position='right') +
        theme_bw() +
        theme(axis.title = element_text(size=32, family='serif'),
              axis.text = element_text(size=18, family='serif'),
              legend.title=element_text(size=18, family='serif'), 
              legend.text=element_text(size=12, family='serif'),
              legend.position='left',
              panel.background=element_blank(),
              panel.border=element_blank(),
              panel.grid.major=element_blank(),
              panel.grid.minor=element_blank(),
              plot.background=element_blank()
        ) +
        labs(fill = "Scaled Value")
```

### Modalities Heatmap in {ggplot2}, Viridis colors, Scaled by max value, and Computationally-sorted 

This final plot uses principal component analysis (PCA) through the {seriation} package to sort and cluster the heatmap both by modalities and themes. 

```{r, echo=FALSE, fig.width=16, fig.height=9}
modalities_scaled_ordered <-
        modalities_data %>%
        rownames_to_column(var = "theme") %>%
        mutate(across(Offline:`Social Media`, rescale_modalities)) %>%
        column_to_rownames('theme') %>%
        as.matrix() %>%
        seriate()

new_levels_scaled_theme <- row.names(modalities_matrix)[get_order(modalities_scaled_ordered, dim=1)]
new_levels_scaled_modality <- colnames(modalities_matrix)[get_order(modalities_scaled_ordered, dim=2)]

modalities_to_plot_ordered <- 
        modalities_to_plot_unordered
modalities_to_plot_ordered$theme <- 
        factor(modalities_to_plot_ordered$theme, levels = new_levels_scaled_theme)
modalities_to_plot_ordered$modality <- 
        factor(modalities_to_plot_ordered$modality, levels = new_levels_scaled_modality)

ggplot(modalities_to_plot_ordered, aes(x = modality, y = theme)) +
        geom_tile(aes(fill = value), color = 'white', size = 0.5) + 
        scale_fill_viridis(discrete = FALSE) +
        xlab("Modality") +
        ylab("Theme") +
        scale_y_discrete(position='right') +
        theme_bw() +
        theme(axis.title = element_text(size=32, family='serif'),
              axis.text = element_text(size=18, family='serif'),
              legend.title=element_text(size=18, family='serif'), 
              legend.text=element_text(size=12, family='serif'),
              legend.position='left',
              panel.background=element_blank(),
              panel.border=element_blank(),
              panel.grid.major=element_blank(),
              panel.grid.minor=element_blank(),
              plot.background=element_blank()
        ) +
        labs(fill = "Scaled Value")
```

```{r, include=FALSE, eval=FALSE}
ggsave("dissertation-modalities-heatmap-sorted.png", width = 16, height = 9)
```

### Themes Heatmap in {ggplot2}, Viridis colors, Scaled by percent presence, and Computationally-sorted 

```{r, echo=FALSE, fig.width=16, fig.height=9}
modalities_quote_count <- c(102, 12,  103)
modalities_scaled <- modalities_matrix
for(i in 1:3) {
       modalities_scaled[, i] <- 
               100 * modalities_scaled[, i] / modalities_quote_count[i]
}

modalities_scaled_ordered2 <- seriate(modalities_scaled)

new_levels_scaled_theme2 <- row.names(modalities_matrix)[get_order(modalities_scaled_ordered2, dim=1)]
new_levels_scaled_modality2 <- colnames(modalities_matrix)[get_order(modalities_scaled_ordered2, dim=2)]

modalities_to_plot_ordered2 <- 
        modalities_scaled %>%
        as_tibble() %>%
        mutate(theme = row.names(modalities_scaled)) %>%
        select(theme, everything()) %>%
        pivot_longer(cols = Offline:`Social Media`, 
                     names_to = "modality")
modalities_to_plot_ordered2$theme <- 
        factor(modalities_to_plot_ordered2$theme, levels = new_levels_scaled_theme2)
modalities_to_plot_ordered2$modality <- 
        factor(modalities_to_plot_ordered2$modality, levels = new_levels_scaled_modality2)

ggplot(modalities_to_plot_ordered2, aes(x = modality, y = theme)) +
        geom_tile(aes(fill = value), color = 'white', size = 0.5) + 
        scale_fill_viridis(rescaler = function(x, to = c(0, 1), from = NULL) {
                ifelse(x < 60, 
                       scales::rescale(x,
                                       to = to,
                                       from = c(min(x, na.rm = TRUE), 60)),
                       1)
        }
        ) +
        xlab("Modality") +
        scale_x_discrete(position='top') +
        ylab("Thematic Code") +
        scale_y_discrete(position='right') +
        theme_bw() +
        theme(axis.title = element_text(size=32, family='serif'),
              axis.text = element_text(size=18, family='serif'),
              legend.title = element_text(size=18, family='serif'), 
              legend.text = element_text(size=12, family='serif'),
              legend.position='bottom',
              legend.background = element_blank(),
              legend.box.background = element_rect(colour='black'),
              panel.background = element_blank(),
              panel.border = element_blank(),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              plot.background = element_blank()
        ) +
        labs(fill = "Percentage of Quotations")
```

```{r, include=FALSE, eval=FALSE}
ggsave("dissertation-modalities-heatmap-sorted2.png", width = 16, height = 9)
```
